### 2015-03-02 src/4_OnlineLearning.tex
Bias variance on lambda ready

\subsection{The bias-variance tradeoff}

@@

\begin{align}
\mathrm{E}\Big[\big(y - \hat{f}(x)\big)^2\Big]
 & = \mathrm{Bias}(\hat{f}(x))^2 + \mathrm{Var}(\hat{f}(x)) + \sigma^2
\end{align}
\noindent where:
\begin{align*}
 \mathrm{Bias}(\hat{f}(x)) &= \mathrm{E}\big[\hat{f}(x)\big] - f(x) \\
 \mathrm{Var}(\hat{f}(x)) &= \mathrm{E}\Big[ \big( \hat{f}(x) - \mathrm{E}[\hat{f}(x)] \big)^2 \Big] 
\end{align*}

@@

@@

%\subsection{The bias-variance tradeoff}
%
%To determine parameter $\lambda$ is crucial for ridge regression since it could
%reduce the expected prediction error by reducing variance (bias-variance trafeoff) considering a biased
%estimator. It is known that the prediction error is
%obtained as:
%
%\begin{equation}
% \label{eq:prederror}
%E[(B-\hat{f}(\mathbf{X}))^2] = \sigma^2 + Bias(\hat{f}(\mathbf{X}))^2 + Var(\hat{f}(\mathbf{X}))
%\end{equation}
%
%\noindent where $\hat{f}(\mathbf{X})=\mathbf{AX}(\lambda)$ and
%$f(\mathbf{X})=\mathbf{AX}$.
%
%
%
%The bias of OLS is:
%
%\begin{eqnarray*}
%Bias(\hat{\mathbf{X}}) &=& E[\hat{\mathbf{X}}] - \mathbf{X} \\
%&=& E[ (\mathbf{A}^\top \mathbf{A})^{-1}\mathbf{A}^\top \mathbf{B}] - \mathbf{X} \\
%&=& E[ (\mathbf{A}^\top \mathbf{A})^{-1}\mathbf{A}^\top (\mathbf{AX})] - \mathbf{X}  \\
%&=& \mathbf{X}  - \mathbf{X}  \\
%&=&  0
%\end{eqnarray*}
%
%
%The bias of ridge regression when $\mathbf{A A^\top}$ is non-singular
%can be obtained expressing ridge regression solution
%$\mathbf{\lambda}$ in terms of OLS solution $\hat{\mathbf{X}}$:
%
%\begin{eqnarray*}
%\mathbf{X}(\lambda) &=&( \mathbf{A}^\top \mathbf{A} + \lambda \mathbb{I})^{-1}\mathbf{B} \\
%&=& (\mathbb{I} + \lambda (\mathbf{A}^\top \mathbf{A})^{-1})^{-1} (\mathbf{A}^\top \mathbf{A})^{-1}\mathbf{A}^\top \mathbf{B} \\
%&=&  (\mathbb{I} + \lambda (\mathbf{A}^\top \mathbf{A})^{-1})^{-1}  \hat{\mathbf{X}} \\
%&=& \mathbf{W} \hat{\mathbf{X}} 
%\end{eqnarray*}
%
%\noindent where $\mathbf{W}  = (\mathbb{I} + \lambda (\mathbf{A}^\top
%\mathbf{A})^{-1})^{-1}  $ it is defined for simplicity. Ridge
%regression bias is then obtained as:
%
%\begin{eqnarray*}
%Bias(\mathbf{X}(\lambda)) &=& E[\mathbf{X}(\lambda)] - \mathbf{X} \\
%&=& E[\mathbf{W}\hat{\mathbf{X}}] - \mathbf{X} \\
%&=&  \mathbf{W} \mathbf{X} - \mathbf{X} \neq 0 
%\end{eqnarray*}
%
%
%
%The variance of OLS is:
%
%\begin{equation*}
%Var(\hat{\mathbf{X}}) = \sigma^2 (\mathbf{A}^\top \mathbf{A} )^{-1}
%\end{equation*}
%
%\noindent and the variance of ridge regression is:
%
%\begin{eqnarray*}
%Var(\mathbf{X}(\lambda)) &=& Var(\mathbf{W}\hat{\mathbf{X}}) \\ 
%&=& \mathbf{W}Var(\hat{\mathbf{X}})\mathbf{W}^\top \\
%&=& \sigma^2 \mathbf{W}(\mathbf{A}^\top \mathbf{A} )^{-1}\mathbf{W}^\top
%\end{eqnarray*}
%
%
%\begin{figure}[h!]
%\includegraphics[width=0.8\linewidth]{plots/biasvariance}
%\caption{Bias-variance tradeoff}
%\label{fig:biasvariance}
%\end{figure}
%
%The figure~\ref{fig:biasvariance} shows the bias-variance tradeoff
%given by equation~(\ref{eq:prederror}). Ridge regression shows an
%increasing bias and a decreasing variance. Despite OLS has zero bias,
%its variance is greater than ridge for small values of $\lambda$.
%It can be shown that, in terms of prediction error, ridge (black line)
%is lower than OLS (dotted line)~\cite{hoerl+kennard1970}.
%
%
%

>>>>

\subsection{The bias-variance tradeoff} \label{sec:biasvar}

@@

\begin{align}
\label{eq:bvtrade}
\mathrm{E}\Big[\big(y - \hat{f}(x)\big)^2\Big]
 & = \mathrm{Bias}(\hat{f}(x))^2 + \mathrm{Var}(\hat{f}(x)) + \sigma^2
\end{align}
\noindent where:
\begin{align*}
 \mathrm{Bias}(\hat{f}(x)) &= \mathrm{E}\big[\hat{f}(x)\big] - f(x) \\
 \mathrm{Var}(\hat{f}(x)) &= \mathrm{E}\Big[ \big( \hat{f}(x) - \mathrm{E}[\hat{f}(x)] \big)^2 \Big] 
\end{align*}

@@

\subsection{The Lambda parameter}

@@

One of the ways to determine parameter $\lambda$ is using the bias-variance tradeoff (see section \ref{sec:biasvar}). This parameter is crucial for ridge regression since it could
reduce the expected prediction error by reducing variance, considering a biased
estimator. 
It is known that the prediction error can be express as a decomposition between bias and variance. The prediction error is
obtained as:

\begin{equation}
 \label{eq:prederror}
E[(Y-\hat{f}(\mathbf{X}))^2] = \sigma^2 + Bias(\hat{f}(\mathbf{X}))^2 + Var(\hat{f}(\mathbf{X}))
\end{equation}

\noindent where $\hat{f}(\mathbf{X})=\mathbf{AX}(\lambda)$ for RR and
$\hat{f}(\mathbf{X})=\mathbf{AX}$ in the case of OLS.

For ridge regression we will use the following notation: $\hat{f}(\mathbf{X})=\mathbf{AX}(\lambda)$ and
for OLS $f(\mathbf{X})=\mathbf{AX}$. The solution of OLS is well known $\hat{\mathbf{X}}=(\mathbf{A}^\top \mathbf{A})^{-1}\mathbf{A}^\top \mathbf{Y}$, and its bias is:

\begin{eqnarray*}
Bias(\hat{\mathbf{X}}) &=& E[\hat{\mathbf{X}}] - \mathbf{X} \\
&=& E[ (\mathbf{A}^\top \mathbf{A})^{-1}\mathbf{A}^\top \mathbf{Y}] - \mathbf{X} \\
&=& E[ (\mathbf{A}^\top \mathbf{A})^{-1}\mathbf{A}^\top (\mathbf{AX})] - \mathbf{X}  \\
&=& \mathbf{X}  - \mathbf{X}  \\
&=&  0
\end{eqnarray*}

The bias of ridge regression when $\mathbf{A A^\top}$ is non-singular
can be obtained expressing ridge regression solution
$\mathbf{\lambda}$ in terms of OLS solution $\hat{\mathbf{X}}$:

\begin{eqnarray*}
\mathbf{X}(\lambda) &=&( \mathbf{A}^\top \mathbf{A} + \lambda \mathbb{I})^{-1}\mathbf{Y} \\
&=& (\mathbb{I} + \lambda (\mathbf{A}^\top \mathbf{A})^{-1})^{-1} (\mathbf{A}^\top \mathbf{A})^{-1}\mathbf{A}^\top \mathbf{Y} \\
&=&  (\mathbb{I} + \lambda (\mathbf{A}^\top \mathbf{A})^{-1})^{-1}  \hat{\mathbf{X}} \\
&=& \mathbf{W} \hat{\mathbf{X}} 
\end{eqnarray*}

\noindent where $\mathbf{W}  = (\mathbb{I} + \lambda (\mathbf{A}^\top
\mathbf{A})^{-1})^{-1}  $ it is defined for simplicity. Ridge
regression bias is then obtained as:

\begin{eqnarray*}
Bias(\mathbf{X}(\lambda)) &=& E[\mathbf{X}(\lambda)] - \mathbf{X} \\
&=& E[\mathbf{W}\hat{\mathbf{X}}] - \mathbf{X} \\
&=&  \mathbf{W} \mathbf{X} - \mathbf{X} \neq 0 
\end{eqnarray*}

\begin{figure}[h!]
\includegraphics[width=\linewidth]{img/biasvariance}
\caption{Bias-variance tradeoff depending on lambda}
\label{fig:biasvariance}
\end{figure}

The variance of OLS is:

\begin{equation*}
Var(\hat{\mathbf{X}}) = \sigma^2 (\mathbf{A}^\top \mathbf{A} )^{-1}
\end{equation*}

\noindent and the variance of ridge regression is:

\begin{eqnarray*}
Var(\mathbf{X}(\lambda)) &=& Var(\mathbf{W}\hat{\mathbf{X}}) \\ 
&=& \mathbf{W}Var(\hat{\mathbf{X}})\mathbf{W}^\top \\
&=& \sigma^2 \mathbf{W}(\mathbf{A}^\top \mathbf{A} )^{-1}\mathbf{W}^\top
\end{eqnarray*}

The figure~\ref{fig:biasvariance} shows the bias-variance tradeoff
given by equation~(\ref{eq:prederror}). Ridge regression shows an
increasing bias and a decreasing variance. Despite OLS has zero bias,
its variance is greater than ridge for small values of $\lambda$.
It can be shown that, in terms of prediction error, ridge (black line)
is lower than OLS (dotted line)~\cite{hoerl1970}.
